---
title: "Part 1 - Tree growth calculations"
author: "mcgrathjes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1 - Importing files, data wrangling, mathematical operations, plots and saving code on GitHub

The file “growth_data.csv” contains measurements for tree circumference growing at two sites, control site and treatment site which were planted 20 years ago.

### 6. Import this csv file into an R object. What are the column names?
Using raw data file from public repository previously listed, the "growth_data" csv file is imported to the project for analysis. 

The column names are listed as follows:
```{r}
URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
 
growth <- read.csv(URL)
cat(colnames(growth), sep = "\n")

```
### 7. Calculate the mean and standard deviation of tree circumference at the start and end of the study at both sites.
Using column 3 "Circumf_2005_cm" as the **start of the study**, and column 6 "Circumf_2020_cm" as the **end of the study**, we calculate the mean and standard deviation grouped by the two study sites: "northeast" and "southwest".

```{r}


```


### 8. Make a box plot of tree circumference at the start and end of the study at both sites.


### 9. Calculate the mean growth over the last 10 years at each site.


### 10. Use the t.test to estimate the p-value that the 10 year growth is different at the two sites.